---
title: "RepData_PeerAssessment1"
author: "Rafael Mendez"
date: "August 7, 2015"
output: html_document
---

Before Anything: Prepare Global Enviornment 
```{r tidy=FALSE}
library(knitr)
library(data.table)
library(Hmisc)
library(ggplot2)
library(lattice)
opts_chunk$set(echo = TRUE, results = 'hold')       

```

1. Loading and preprocessing the data

    A) Load in the activity.csv Data and Read into a dataframe
```{r tidy=FALSE}
activitydata<-read.csv("activity.csv", header = TRUE, sep = ",", 
colClasses = c("numeric","character","numeric"))
```

    B) Check the given data frame with str()
```{r}
    str(activitydata)
```

    C) Tidy the data frames date and interval field 
```{r}
    activitydata$date <- as.Date(activitydata$date)
```
    

2. What is mean total number of steps taken per day?

    A) Calculate the sum of the total steps per day
```{r}
    Total_steps_day <- aggregate(steps ~ date, data = activitydata, sum, na.rm = TRUE)
    colnames(Total_steps_day ) <- c("date","steps")
    head(Total_steps_day,10)
```
    
    B) Histogram: Total Number of Steps per day with missing values included  
```{r tidy=FALSE}
    ggplot(Total_steps_day, aes_string(x = "steps")) +
    geom_histogram(col= "black",bindwidth = 1000, fill ="orange") +
    ggtitle("Activity Histogram: Steps per Day (Missing Values Removed") + 
    xlab("Number of Steps per Day") + 
    ylab("Number of times in day(Count)")
```

    C) Report the mean and median of the total number of steps taken per day
```{r}
    Total_steps_mean <- mean(Total_steps_day$steps, na.rm = TRUE)
    Total_steps_median <- median(Total_steps_day$steps, na.rm = TRUE)
    Total_steps_mean    
    Total_steps_median
```

3. What is the average daily activity pattern?

    A) TimeSeries Plot: Average Daily Activity Pattern of Steps Per Day
```{r tidy=FALSE}
    #Can use tapply or aggregate, choose aggregate to make into a data frame 
    #which is easier to use than a matrix. 
    
    #Steps per interval
    time_series <- aggregate(activitydata$steps, by = list(interval = activitydata$interval), 
    FUN=mean, na.rm = TRUE)
    colnames(time_series) <- c("interval", "steps") #steps is x if this is not done
    
    #Use xyplot for bivariate time-seris plots
    xyplot(steps~interval, data=time_series, type="l", grid=TRUE, col ="red", xlab="5-min Intervals", 
    ylab="Average Steps on all Days", main="Average Daily Activity Patterns:Steps", aspect="fill", 
    size=5, border=TRUE)
```
    
    B) Report which 5-minute interval has the max average of steps
```{r}
    #Find the row which has the max number of steps use [] to access elements in data frame and the 'which.max()' function to find index
    max_avg_interval <- time_series[which.max(time_series$steps),]
    max_avg_interval #shows what interval has max number of steps and what steps 
```

4. Imputing missing values - Missing days may Introduce Bias. Show this

    A) Report Number of Missing Values 
```{r}
    #Report the total number of missing values in the dataset 
    missingvalue <- sum(is.na(activitydata$steps))
    missingvalue
```

    B) Make new dataset with NA filled in
       Solution:Replace missing values with average steps per interval from time_seris plot
```{r}
Filled_Data <- activitydata
Filled_Data$steps <- impute(activitydata$steps, fun=mean)
head(Filled_Data,10)           #Compare Data without NA
sum(is.na(Filled_Data$steps))  #Check if any NA left in Dataset
```

    C) Histogram: Total Number of Steps per day with missing values filled (replaced) 
```{r tidy=FALSE}
    steps_day_NA <- tapply(Filled_Data$steps, Filled_Data$date, sum)
    qplot(steps_day_NA, xlab='Total steps per day (Imputed)', ylab='Frequency using binwith 500', binwidth=500)
```
    
    D) Report the Mean and Median Total: Total Steps Per Day
```{r}
Total_steps_mean_NA <- mean(steps_day_NA)
Total_steps_median_NA <- median(steps_day_NA)
Total_steps_mean_NA
Total_steps_median_NA
```

    E) Report if these values differ from the estimates from having missing values included in the first dataset and report the impact of imputting missing data on estimates of the total daily number of steps
    
    Result: Yes the values in means and median changed. Imputing missing values have more data, thus the     greater mean and median.
```{r}
#Before Missing Values are Imputted
#Mean
Total_steps_mean
#Median
Total_steps_median

#Imputted Missing Values
#Mean
Total_steps_mean_NA
#Median
Total_steps_median_NA

```

5. Are there differences in activity patterns between weekdays and weekends?

    A) Create a new factor variable in the dataset with two levels- "weekday" and "weekend" incdicating
    whether a given date is a weekday or weekend day
    
```{r}
# Compute the weekdays from the date attribute
Filled_Data <- data.frame(date=Filled_Data$date, 
                           weekday=tolower(weekdays(Filled_Data$date)), 
                           steps=Filled_Data$steps, 
                           interval=Filled_Data$interval)

# Create field daytype to identify weekdend and weekdays
Filled_Data <- cbind(Filled_Data, 
                      daytype=ifelse(Filled_Data$weekday == "saturday" |  #test is saturday or sunda 
                                     Filled_Data$weekday == "sunday", 
                                     "weekend",                           #if either then weekend
                                     "weekday"))                          #else is a weekday

# Create the final data.frame
Filled_Data_Weeks <- data.frame(date=Filled_Data$date, 
                       weekday=Filled_Data$weekday, 
                       daytype=Filled_Data$daytype, 
                       interval=Filled_Data$interval,
                       steps=Filled_Data$steps)
```
    
    B) Panel Plot Containing TIme_Series of 5-minute interval (x-axis) and average number of steps 
    taken, averaged accross all weekday days or weekend days (y-axis). 
    
```{r}
#Find Avg Number of steps taken averaged accross all weekday days or weenend days 
avg_steps_week <- aggregate(Filled_Data_Weeks$steps, 
                            by=list(Filled_Data_Weeks$daytype, 
                            Filled_Data_Weeks$weekday, Filled_Data_Weeks$interval), mean)
colnames(avg_steps_week) <- c("daytype", "weekday", "interval", "Avg_steps")
```

Create the Time_Series Plot 
```{r}
xyplot(Avg_steps ~ interval | daytype, data = avg_steps_week,
       type = "l",
       lwd = 1,
       xlab= "Interval",
       ylab = "Number of Steps Per Interval",
       layout=c(1,2))

```